Exploring the ages of rotating stars using galactic dynamics: a novel approach to calibrating gyrochronology

### 2019-08-30 paper/intro.tex
adding some discussion about ages on the CMD and adding structural organization of paper

In this paper, we tested the \citet{angus2019} gyrochronology relation, a
separable relation, calibrated using the period-color relation of Praesepe (in
\gaia\ \gcolor\ color) and the period-age relation of Praesepe and the Sun.
The large number of Praesepe members with precise rotation periods from the
\ktwo\ mission \citep{douglas2017, rebull2017}, spanning spectral types F
through early M, makes it the ideal cluster for calibrating the period-color
relation of stars at 650 Myrs.
% This relation was calibrated by fitting a 5th-order polynomial to the relation
% between (log) rotation period and (log) \gaia\ \gcolor\ color for around 800
% members of the Praesepe cluster, and a straight line in (log) age to Praesepe
% and the Sun.
This relation accurately describes the rotation periods of F and G stars up to
around 2.5 Gyr (the age of NGC 6819 -- the oldest cluster with available
rotation periods), but over-predicts the rotation periods of K dwarfs in the
1.1 Gyr NGC 6811 cluster.
Very few reliable age estimates exist for K dwarfs with rotation periods older
than 1.1 Gyr, or between the ages of $\sim$ 800 Myr--1.1 Gyr so the rotational
evolution of middle-aged K dwarfs is a mystery\footnote{The oldest M dwarfs
with reliable ages and rotation periods are members of Praesepe ($\sim$ 650
Myrs), so even less is known about their rotational evolution.}.
% In this paper, we shed light on the rotational evolution of old K dwarfs using
% population-based dynamical ages.
% Unfortunately, open clusters with rotation period measurements are mostly
% young -- currently the oldest is 2.5 Gyr (NGC 6819).
% Without rotation periods for precisely dated old stars, it is extremely
% difficult to calibrate the relationship between rotation period and color at
% old ages.
For this reason, we used a population-based stellar age indicator, velocity
dispersion, to investigate the period-color relations at old ages in the
field.

@@

>>>>

In this paper, we tested the \citet{angus2019} gyrochronology relation, a
separable relation, calibrated using the period-color relation of Praesepe (in
\gaia\ \gcolor\ color) and the period-age relation of Praesepe and the Sun.
The large number of Praesepe members with precise rotation periods from the
\ktwo\ mission \citep{douglas2017, rebull2017}, spanning spectral types F
through early M, makes it a good cluster for calibrating the period-color
relation of stars at 650 Myrs.
% This relation was calibrated by fitting a 5th-order polynomial to the relation
% between (log) rotation period and (log) \gaia\ \gcolor\ color for around 800
% members of the Praesepe cluster, and a straight line in (log) age to Praesepe
% and the Sun.
This relation accurately describes the rotation periods of F and G stars up to
around 2.5 Gyr (the age of NGC 6819 -- the oldest cluster with available
rotation periods), but over-predicts the rotation periods of K dwarfs in the
1.1 Gyr NGC 6811 cluster.
Very few reliable age estimates exist for K dwarfs with rotation periods older
than 1.1 Gyr, or between the ages of $\sim$ 800 Myr--1.1 Gyr so the rotational
evolution of middle-aged K dwarfs is difficult to characterize\footnote{The
oldest M dwarfs with reliable ages and rotation periods are members of
Praesepe ($\sim$ 650 Myrs), so even less is known about their rotational
evolution.}.

The gyrochronal ages of the stars in our sample were calculated with the
\citep{angus2019} gyrochronology model, using de-reddened \Gaia\ \gcolor\
color and rotation periods reported by \mct.
These rotational ages are shown on the CMD in figure \ref{fig:age_gradient}.
The stars with old rotational ages, plotted in yellow hues, predominantly lie
along the upper edge of the MS, where stellar evolution models predict old
stars to be, however the majority of stars with old rotational ages are bluer
than \gcolor\ $\sim$ 1.5 dex.
This provides a first hint that the \citet{angus2019} gyrochronology relation
under-predicts the ages of low-mass stars.
There is no reason to expect the oldest stars in this sample to be the bluer
ones: M dwarfs are, on average, older than K dwarfs and are expected to remain
active for longer, so should therefore have measurable rotation periods at
older ages.
Since the \citet{angus2019} gyrochronology model, which is based on the
period-color relation of Praesepe, does not predict old ages for any M dwarfs,
it is probably under-predicting the ages of some of these low-mass stars.
It is therefore likely that old M dwarfs rotate more rapidly than this
Praesepe-based model predicts.
In what follows we use a population-based stellar age indicator, velocity
dispersion, to investigate the period-color relations at old ages in the
field.

% In this paper, we shed light on the rotational evolution of old K dwarfs using
% population-based dynamical ages.
% Unfortunately, open clusters with rotation period measurements are mostly
% young -- currently the oldest is 2.5 Gyr (NGC 6819).
% Without rotation periods for precisely dated old stars, it is extremely
% difficult to calibrate the relationship between rotation period and color at
% old ages.
% For this reason, we used a population-based stellar age indicator, velocity
% dispersion, to investigate the period-color relations at old ages in the
% field.

@@

This paper is laid out as follows: in section \ref{sec:the_data} we describe
our sample selection process and the methods used to calculated stellar
velocities.
In section \ref{sec:results} we use stellar kinematics to investigate the
relationship between rotation period, age and color/\teff\ in field stars.
In section \ref{sec:mass-dependent-heating} we show that neither
mass-dependent heating nor the selection function is likely to have a strong
affect on our sample.
Finally, in section \ref{sec:period_gap} we discuss the implications of our
results.
